Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2017-04-29 FreeLogic/2017-Journal/article.tex
update

\def\FFOL{\mathit{FFOL}}
\def\V{\mathit{V}}
\def\F{\mathit{F}}
\def\P{\mathit{P}}
\def\True{\mathit{True}}
\def\False{\mathit{False}}
\def\Impl{\rightarrow}
\def\Descr{\rotatebox[origin=C]{180}{$\iota$}}
%\def\Descr{\rotate[c]{180}{\iota}}

@@

\section{Introduction}
\label{intro}
Partiality and undefinedness are prominent challenges in various areas
of mathematics and computer science.  Unfortunately, however, modern
proof assistant systems and automated theorem provers based on
traditional classical or intuitionistic logics provide rather
inadequate support for these challenge concepts.  Free logic \cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} offers a
theoretically appealing solution, but is has been considered as rather
unsuited towards practical utilisation.

@@

Our exploration has been significantly supported by series of
experiments in which automated reasoning tools have been called from
within the proof assistant Isabelle/HOL via the
Sledgehammer tool. Moreover, we have obtained
very useful feedback at various stages from the model finder Nitpick
\cite{Nitpick} saving us from making several mistakes.

@@

For a detailed discussion of the syntax, semantics and automation
of  HOL we refer to the literature (see e.g.
\cite{AndrewsSEP,B5,J6} and the
references therein). Below we introduce a variant of \HOL with primitive equality
and description.

\begin{definition}[Types]\label{holtypes}
  The set $\hol{T}$ of simple types freely generated from a set of
  basic types $\{\hol{o}, \hol{i}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans and 
  $\hol{i}$ is the type of individuals. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{definition}

\begin{definition}[Syntax of $\HOL$]\label{holgrammar}
The language of higher-order logic $\HOL$ with primitive equality and
description is defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \S1.4 of Andrews's overview
  article~\cite{AndrewsSEP}. The choice here closely reflects the
  set of primitive connectives as chosen in higher-order automated theorem provers such
  as LEO-II \cite{LEO}, Leo-III \cite{Leo-III} and Satallax \cite{Satallax}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
formulas.  The logical connectives of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{=_{\alpha\ar\alpha\ar  o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (where $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{definition}

@@

\begin{definition}[Model for \HOL]\label{holmodel}
  A \emph{model} for \HOL is a tuple $\hol{M}=\hol{\langle D, I \rangle}$, where
  $\hol{D}$ is a frame, and $\hol{I}$ is a family of typed interpretation
  functions mapping constant symbols $\hol{p_\alpha}$ to appropriate
  elements of $\hol{D_\alpha}$, called the \emph{denotation of $\hol{p_\alpha}$}
  (the logical connectives $\hol{\neg}$, $\hol{\vee}$, and $\hol{\forall}$ are always
  given the standard denotations, see below).  Moreover, we assume that the domains
  $\hol{D_{\alpha\ar\alpha\ar o}}$ contain the respective identity relations.
\end{definition}

\begin{definition}[Evaluation Function for \HOL]\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOL term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle D, I \rangle}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ 
\item $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|s_\alpha = t_\alpha\|^{M,g}} = {\True}$ if and only if $\hol{\|s_\alpha\|^{M,g}} = \hol{\|t_\alpha\|^{M,g}} $
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = {\True}$ if and only if $\hol{\|s_o\|^{M,g}} = {\False}$
\item $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  {\True}$ if and only if\, $\hol{\|s_o\|^{M,g}} = {\True}$ or $\hol{\|t_o\|^{M,g}}
  = {\True}$
\item \sloppy $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = {\True}$ if and only if for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = {\False}$
\item $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{d}$ if there exists a unique $\hol{d}\in \hol{D_{\alpha}}$ such that
  $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = {\True}$, 
otherwise  $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{e}$ for an arbitrary element $\hol{e}\in \hol{D_{\alpha}}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{definition}

@@

\section{Shallow Semantical Embedding of $\FFOL$ in $\HOL$}
We now present a shallow embedding of $\FFOL$ in $\HOL$ by identifying
the language constructs of $\FFOL$ with corresponding terms of $\HOL$.
In this embedding the raw domain $D$ of a $\FFOL$ model is
identified with the domain of individuals $\hol{D_i}$ in a corresponding $\HOL$ model.  The subdomain $E$ of
existing objects is characterized in the embedding by a $\HOL$
predicate $\hol{\ex}$ of type $\hol{ i\ar o}$. Hence, we assume in the
remainder that a respective uninterpreted constant symbol
$\hol{\ex_{i\ar o}}$
is given in the signature of $\HOL$. Moreover, we assume that an
uninterpreted constant symbol $\hol{\star}$ of type $\hol{i}$ be in the
signature of $\HOL$. Finally, we assume that $\hol{\|E \star_i\|^{M,g}=F}$ for
all $\hol{M,g}$, i.e. that the element denoted by $\hol{\star_i}$ is not
an element of the domain of existing objects denoted by
$\hol{\ex_{i\ar o}}$.

@@

The above mapping induces mappings from the sets $\map{V}$, $\map{F}$ and $\map{P}$
of $\FFOL$ to corresponding variables (of type $i$), uninterpreted function
symbols and uninterpreted predicate symbols in \HOL.

@@

The above results enable the employment of any theorem prover that
supports $\HOL$ with definite description to reason with $\FFOL$,
including TPTP THF \cite{J22} compliant systems such as Satallax,
Nitpick, LEO-II and Leo-III. Alternatively, this theory can be encoded
in interactive proof assistants, such as Isabelle/HOL. <
  % \footnote{An example of respective abstract level proof
  % tactics in the context of a semantical embedding of an
  % hyperintensional higher-order modal logic in $\HOL$ can be found at:
  % \url{https://github.com/ekpyron/TAO}; in this document a
  % computer-formalization of Ed Zalta's Principia
  % Logico-Metaphysica \url{https://mally.stanford.edu/principia.pdf} is presented.}

Properties of $\FFOL$ could now be experimentally explored with
automated reasoning tools for $\HOL$. We have conducted such
experiments in 
an earlier paper \cite{ICMS}. These experiments confirm the
illustrating examples discussed in Scott's paper \cite{Scott67}.

\section{Exploring Axioms Systems for Category Theory}
In an experimental theory exploration study, in which we utilized the
free logic reasoning farmework from above, 
we have shown how Scott's
\cite{Scott79} axiom system for category theory can be derived from a
notion of partial monoids. These axioms systems are presented in Table \ref{axioms-sets-1}.

The stepwise evolution has been described in detail
in\cite{ArXiv}. Below we summarize these experiments.  However, first
we describe some basic modeling decisions for the technical encoding
in Isabelle/HOL. The sources of our experiments, as discussed in
\cite{ArXiv}, are available at
\url{http://christoph-benzmueller.de/papers/2017-arXiv.zip}. Note that
in these sources we added neither the definite description operator
nor the designated undefined object $\hol{\star}$, since both were not
required in this particular application.

Figure \ref{pic1} displays (parts of) the embedding of $\FFOL$ in
$\HOL$, encoded in Isabelle/HOL, that we have employed in our
experiments. Excerpts of these experiments are illustrated in Fig.~\ref{pic2}. A related Isabelle/HOL encoding of the above theory with
description has been presented and discussed in
\cite{ICMS}.\footnote{In the remainder of this article, and inline
  with our text so far,  we present the formulas of
  $\FFOL$ in non-boldface font. These formulas have been encoded in Isabelle/HOL using the
  abbreviations as introduced in Fig.~\ref{pic1}. In the actual source
  encoding, however, the usage of boldface and non-boldface is
  (for technical reasons) reversed.}

@@

\subsection{Modeling of basic concepts}
Morphisms in the category are modeled as objects in $D$ (respectively,
$\hol{D_i}$). We introduce three partial functions, 
$\dom$ (domain), $\cod$ (codomain), and $\comp$ (morphism composition). 
Partiality of composition is handled exactly as expected: we generally may have 
non-existing compositions $x\comp y$ (i.e. $\neg(\ex(x\comp y))$) for some existing  
morphisms $x$ and $y$ (i.e. $\ex x$ and $\ex y$).

@@

\subsection{Consistency}
The model finder Nitpick confirms consistency for all of the axioms
sets from Table \ref{axioms-sets-1}. For example, when asked to consider at least one defined and one undefined object, then
Nitpick generates for all cases the following model (called $M_1$ in the remainder):
$D=\{i_i,i_2\}$ and $E=\{i_1\}$;  $i_1\comp i_1$ is $i_1$,  and $i_2$
in all other cases; $\cod$ and $\dom$ are identity on $D$. Without constraining the request, Nitpick
generates an even simpler model (called $M_0$ in the remainder):
$D=\{i_i\}$ and $E=\emptyset$; $i_1\comp i_1$ is $i_1$; $\cod$ and
$\dom$ are identity on $D$. It is trivial to check that these models
indeed confirm the consistency of all axioms sets from Table \ref{axioms-sets-1}.

@@

Axioms Set I generalizes the notion of a monoid by introducing a partial, strict binary 
composition operation $\comp$. The existence of left and right identity elements is 
addressed in the last two axioms. The notions of $\dom$ (Domain) and
$\cod$ (Codomain)
abstract from their common meaning in the context of sets. In category theory we
work with just a single type of objects (the type $\hol{i}$ in our setting) and therefore 
identity morphisms are employed to suitably characterize their
meanings.

@@

%\subsection{Axioms Set II}
Axioms Set II is developed from Axioms Set I by Skolemization of the
existentially quantified variables $i$ and $j$ in axioms $C_i$ and
$D_i$. We can argue semantically that every model of Axioms Set I has
such functions. Hence, we get a conservative extension of Axioms Set
I. This could be done for any theory with an ``$\forall x\lambdot \exists i\lambdot$''-axiom. The strictness axiom $S$ is extended, so
that strictness is now also postulated for the new Skolem functions
$\dom$ and $cod$. Note that the values of Skolem functions
outside $E$ can just be given by the identity function.

@@

\subsection{Remark on the Experiments}
All proofs above and all proofs in the rest of this paper (unless
stated otherwise) have been obtained fully automatically in very
reasonable time (typically just a few seconds) with the
Sledgehammer tool in Isabelle/HOL. This tool interfaces to prominent
first-order automated theorem provers such as CVC4 \cite{CVC4}, Z3
\cite{Z3}, E \cite{E} and Spass \cite{Spass}.  Remotely, also
  provers such as Vampire \cite{Vampire}, or the higher-order provers
  Satallax \cite{Satallax} and LEO-II \cite{LEO} can be reached. For
  example, to prove axiom $E_{iii}$ from Axioms Set II, we have called
  Sledgehammer on all axioms of Axioms Set II. The provers then, via
  Sledgehammer, suggested to call trusted/verified tools in
  Isabelle/HOL with the exactly required dependencies they
  detected. With the provided dependency information the trusted tools
  in Isabelle/HOL were then able to reconstruct the external proofs on
  their own.  This way we obtain a verification of our claims in
  Isabelle/HOL, in which all the proofs have nevertheless been
  contributed by automated theorem provers. However, there are still
  a few calls of the \textsc{smt}-tactic contained in our document, 
   which we found are not entirely trivial to replace. In this sense, our
   Isabelle/HOL verification is modulo the correctness of the \textsc{smt} solvers
   CVC4 \cite{CVC4} and Z3 \cite{Z3}.

\subsection{Axioms Set III, IV and V} 
In Axioms Set III the existence  axiom  $E_{ii}$ from Axioms Set II  is simplified by taking advantage of 
  the two new Skolem functions ${\dom}$ and $cod$.

@@

 \subsection{Constricted Inconsistency in Free Logic
   Setting} \label{subsec-freyd-scedrov-1} A main difference in the
 system by Freyd and Scedrov to our Axioms Set V from Table
 \ref{axioms-sets-1} concerns axiom $S3$ respectively $A1$. Namely,
 instead of the non-reflexive existing identity $\simeq$, they use
 Kleene equality $\cong$, cf. definition 1.11 on page 3 of their textbook
 \cite{FreydScedrov90}.\footnote{Def. 1.11 in Freyd Scedrov: ``The
   ordinary equality sign $=$ [i.e., our $\cong$] will be used in the
   symmetric sense, to wit: if either side is defined then so is the
   other and they are equal. \ldots''} The difference seems minor, but
 in our free logic setting it has the effect to cause the mentioned
 constricted inconsistency issue.\footnote{This could perhaps be an
   oversight, or it could indicate that Freyd and Scedrov actually
   mean the Axioms Set discussed in \S\ref{subsec-freyd-scedrov-2}
   below.%  (where the variables in the axioms range over defined objects
   % only). However, in this reading of the axioms of Freyd and Scedrov
   % we had to (re-)introduce explicit strictness conditions to ensure
   % equivalence to the Axioms Set V by Scott.
 }

@@

Moreover, abstract level proof tactics could be provided in proof
assistants to support --- as an alternative to always unfolding the
mapping from $\FFOL$ to $\HOL$ --- intuitive interaction (and even
automation) in $\FFOL$ on top the semantical embedding.

>>>>

\def\FFOL{\entity{FFOL}\xspace}
\def\V{\mathit{V}}
\def\F{\mathit{F}}
\def\P{\mathit{P}}
\def\True{\mathit{True}}
\def\False{\mathit{False}}
\def\Impl{\rightarrow}
\def\Descr{\rotatebox[origin=C]{180}{$\iota$}}
%\def\Descr{\rotate[c]{180}{\iota}}

@@

\section{Introduction}
\label{intro}
Partiality and undefinedness are prominent challenges in various areas
of mathematics and computer science.  Unfortunately, however, modern
proof assistant systems and automated theorem provers based on
traditional classical or intuitionistic logics provide rather
inadequate support for these challenge concepts.  Free logic \cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} offers a
theoretically appealing solution, but is has been considered as rather
unsuited towards practical utilization.

@@

Our exploration has been significantly supported by series of
experiments in which automated reasoning tools have been called from
within the proof assistant Isabelle/HOL via the
Sledgehammer tool. Moreover, we have obtained
very useful feedback at various stages from the model finder Nitpick
\cite{Nitpick}, saving us from making several mistakes.

@@

For a detailed discussion of the syntax, semantics and automation
of  HOL we refer to the literature (see e.g.
\cite{AndrewsSEP,B5,J6} and the
references therein). Below we introduce a variant of \HOL with primitive equality
and definite description.

\begin{definition}[Types]\label{holtypes}
  The set $\hol{T}$ of simple types is freely generated from a set of
  basic types $\{\hol{o}, \hol{i}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans and 
  $\hol{i}$ is the type of individuals. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{definition}

\begin{definition}[Syntax of $\HOL$]\label{holgrammar}
The terms of $\HOL$ with primitive equality and
definite description are defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \S1.4 of Andrews's overview
  article~\cite{AndrewsSEP}. The choice here closely reflects the
  set of primitive connectives as chosen in higher-order automated theorem provers such
  as LEO-II \cite{LEO}, Leo-III \cite{Leo-III} and Satallax \cite{Satallax}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
\emph{formulas}.  The \emph{logical connectives} of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{=_{\alpha\ar\alpha\ar  o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (where $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{definition}

@@

\begin{definition}[Model for \HOL]\label{holmodel}
  A \emph{model (structure)} for \HOL is a tuple $\hol{M}=\hol{\langle D, I \rangle}$, where
  $\hol{D}$ is a frame, and $\hol{I}$ is a family of typed interpretation
  functions mapping constant symbols $\hol{p_\alpha}$ to appropriate
  elements of $\hol{D_\alpha}$, called the \emph{denotation of $\hol{p_\alpha}$}
  (the logical connectives $\hol{\neg}$, $\hol{\vee}$, and $\hol{\forall}$ are always
  given the standard denotations, see below).  Moreover, we assume that the domains
  $\hol{D_{\alpha\ar\alpha\ar o}}$ contain the respective identity relations.
\end{definition}

\begin{definition}[Evaluation Function for \HOL]\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOL term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle D, I \rangle}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ 
\item $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|s_\alpha = t_\alpha\|^{M,g}} = {\True}$ if and only if $\hol{\|s_\alpha\|^{M,g}} = \hol{\|t_\alpha\|^{M,g}} $
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = {\True}$ if and only if $\hol{\|s_o\|^{M,g}} = {\False}$
\item \sloppy $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  {\True}$ if and only if\, $\hol{\|s_o\|^{M,g}} = {\True}$ or $\hol{\|t_o\|^{M,g}}
  = {\True}$
\item \sloppy $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = {\True}$ if and only if for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = {\True}$
\item $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{d}$ if there exists a unique $\hol{d}\in \hol{D_{\alpha}}$ such that
  $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = {\True}$, 
otherwise  $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{e}$ for an arbitrary element $\hol{e}\in \hol{D_{\alpha}}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{definition}

@@

\section{Shallow Semantical Embedding of $\FFOL$ in $\HOL$}
We now present a shallow embedding of $\FFOL$ in $\HOL$ by identifying
the language constructs of $\FFOL$ with corresponding terms of $\HOL$.
In this embedding the raw domain $D$ of a $\FFOL$ model is
identified with the domain of individuals $\hol{D_i}$ in a corresponding $\HOL$ model.  The subdomain $E$ of
existing objects is characterized in the embedding by a $\HOL$
predicate $\hol{\ex}$ of type $\hol{ i\ar o}$. Hence, we assume in the
remainder that a respective uninterpreted constant symbol
$\hol{\ex_{i\ar o}}$
is given in the signature of $\HOL$. Moreover, we assume that an
uninterpreted constant symbol $\hol{\star}$ of type $\hol{i}$ be in the
signature of $\HOL$. Finally, we assume that $\hol{\|E \star_i\|^{M,g}=F}$ for
all $\hol{M,g}$, i.e. that the element denoted by $\hol{\star_i}$ is not
an element of the domain of existing objects denoted by
$\hol{\ex_{i\ar o}}$ (technically this can be achieved by postulating
a respective axiom).

@@

The above mapping induces mappings from the sets $\map{V}$, $\map{F}$ and $\map{P}$
of $\FFOL$ to corresponding variables (of type $i$), uninterpreted function
symbols and uninterpreted predicate symbols in \HOL, respectively.

@@

The above results enable the employment of any theorem prover that
supports $\HOL$ with definite description to reason with $\FFOL$,
including TPTP THF \cite{J22} compliant systems such as Satallax,
Nitpick, LEO-II and Leo-III. Alternatively, this theory can be encoded
in interactive proof assistants, such as Isabelle/HOL. 
  % \footnote{An example of respective abstract level proof
  % tactics in the context of a semantical embedding of an
  % hyperintensional higher-order modal logic in $\HOL$ can be found at:
  % \url{https://github.com/ekpyron/TAO}; in this document a
  % computer-formalization of Ed Zalta's Principia
  % Logico-Metaphysica \url{https://mally.stanford.edu/principia.pdf} is presented.}

The different properties of $\FFOL$ could now be experimentally explored with
automated reasoning tools for $\HOL$. We have conducted such
experiments in 
an earlier paper \cite{ICMS}. These experiments confirm the
illustrating examples discussed in Scott's paper \cite{Scott67}.

\section{Exploring Axioms Systems for Category Theory}
In an experimental theory exploration study, utilizing the
free logic reasoning framework from above, 
we have shown how Scott's
\cite{Scott79} axiom system for category theory can be derived from a
notion of partial monoids. These axioms systems are presented in Table \ref{axioms-sets-1}.

The stepwise evolution has been described in detail in
\cite{ArXiv}. Below we summarize these experiments.  However, first we
describe some basic modeling decisions for the technical encoding in
Isabelle/HOL.

The sources of our experiments, as discussed in
\cite{ArXiv}, are available at
\url{http://christoph-benzmueller.de/papers/2017-arXiv.zip}. Note that
in these sources we added neither the definite description operator
nor the designated undefined object $\hol{\star}$, since both were not
required in this particular application. A related Isabelle/HOL
encoding of the above theory with description has been presented and
discussed in \cite{ICMS}.

Figure \ref{pic1} displays (parts of) the embedding of $\FFOL$ in
$\HOL$, encoded in Isabelle/HOL, that we have employed in our
experiments.\footnote{In the remainder of this article, and inline
  with our text so far,  we present the formulas of
  $\FFOL$ in non-boldface font. These formulas have been encoded in Isabelle/HOL using the
  abbreviations as introduced in Fig.~\ref{pic1}. In the actual source
  encoding, however, the usage of boldface and non-boldface is
  (for technical reasons) reversed.} 
An excerpt of these experiments is shown in Fig.~\ref{pic2}.

@@

\subsection{Modeling of basic concepts}
Morphisms in the category are modeled as objects in $D$ (respectively,
$\hol{D_i}$). We introduce three partial functions, 
$dom$ (domain), $cod$ (codomain), and $\comp$ (morphism composition). 
Partiality of composition is handled exactly as expected: we generally may have 
non-existing compositions $x\comp y$ (i.e. $\neg(\ex(x\comp y))$) for some existing  
morphisms $x$ and $y$ (i.e. $\ex x$ and $\ex y$).

@@

\subsection{Consistency}
The model finder Nitpick confirms consistency for all of the axioms
sets from Table \ref{axioms-sets-1}. For example, when asked to consider at least one defined and one undefined object, then
Nitpick generates for all cases the following model (called $M_1$ in the remainder):
$D=\{i_i,i_2\}$ and $E=\{i_1\}$;  $i_1\comp i_1$ is $i_1$,  and $i_2$
in all other cases; $cod$ and $dom$ are identity on $D$. Without constraining the request, Nitpick
generates an even simpler model (called $M_0$ in the remainder):
$D=\{i_i\}$ and $E=\emptyset$; $i_1\comp i_1$ is $i_1$; $cod$ and
$dom$ are identity on $D$. It is trivial to check that these models
indeed confirm the consistency of all axioms sets from Table \ref{axioms-sets-1}.

@@

Axioms Set I generalizes the notion of a monoid by introducing a partial, strict binary 
composition operation $\comp$. The existence of left and right identity elements is 
addressed in the last two axioms. The notions of $dom$ (domain) and
$cod$ (codomain)
abstract from their common meaning in the context of sets. In category theory we
work with just a single type of objects (the type $\hol{i}$ in our setting) and therefore 
identity morphisms are employed to suitably characterize their
meanings.

@@

%\subsection{Axioms Set II}
Axioms Set II is developed from Axioms Set I by Skolemization of the
existentially quantified variables $i$ and $j$ in axioms $C_i$ and
$D_i$. We can argue semantically that every model of Axioms Set I has
such functions. Hence, we get a conservative extension of Axioms Set
I. This could be done for any theory with an ``$\forall x\lambdot \exists i\lambdot$''-axiom. The strictness axiom $S$ is extended, so
that strictness is now also postulated for the new Skolem functions
$dom$ and $cod$. Note that the values of Skolem functions
outside $E$ can just be given by the identity function.

@@

\subsection{Remark on the Experiments}
All proofs above and all proofs in the rest of this paper (unless
stated otherwise) have been obtained fully automatically in very
reasonable time (typically just a few seconds) with the
Sledgehammer tool in Isabelle/HOL. This tool interfaces to prominent
first-order automated theorem provers such as CVC4 \cite{CVC4}, Z3
\cite{Z3}, E \cite{E} and Spass \cite{Spass}.  Remotely, also
  provers such as Vampire \cite{Vampire}, or the higher-order provers
  Satallax \cite{Satallax} and LEO-II \cite{LEO} can be reached. For
  example, to prove axiom $E_{iii}$ from Axioms Set II, we have called
  Sledgehammer on all axioms of Axioms Set II. The provers then, via
  Sledgehammer, suggested to call trusted/verified tools in
  Isabelle/HOL with the exactly required dependencies they
  detected, in this case $C_{ii}$, $D_{ii}$, $E_{ii}$ and $S_{ii}$. With the provided dependency information the trusted tools
  in Isabelle/HOL were then able to reconstruct the external proofs on
  their own.  This way we obtain a verification of our claims in
  Isabelle/HOL, in which all the proofs have nevertheless been
  contributed by automated theorem provers. However, there are still
  a few calls of the \textsc{smt}-tactic contained in our document, 
   which we found are not entirely trivial to replace. In this sense, our
   Isabelle/HOL verification is modulo the correctness of the \textsc{smt} solvers
   CVC4 \cite{CVC4} and Z3 \cite{Z3}.

\subsection{Axioms Sets III, IV and V} 
In Axioms Set III the existence  axiom  $E_{ii}$ from Axioms Set II  is simplified by taking advantage of 
  the two new Skolem functions ${\dom}$ and $cod$.

@@

 \subsection{Constricted Inconsistency in Free Logic
   Setting} \label{subsec-freyd-scedrov-1} A main difference in the
 system by Freyd and Scedrov to our Axioms Set V from Table
 \ref{axioms-sets-1} concerns axiom $S3$, respectively $A1$. Namely,
 instead of the non-reflexive existing identity $\simeq$, they use
 Kleene equality $\cong$, cf. definition 1.11 on page 3 of their textbook
 \cite{FreydScedrov90}.\footnote{Def. 1.11 in Freyd Scedrov: ``The
   ordinary equality sign $=$ [i.e., our $\cong$] will be used in the
   symmetric sense, to wit: if either side is defined then so is the
   other and they are equal. \ldots''} The difference seems minor, but
 in our free logic setting it has the effect to cause the mentioned
 constricted inconsistency issue.\footnote{This could perhaps be an
   oversight, or it could indicate that Freyd and Scedrov actually
   mean the Axioms Set discussed in \S\ref{subsec-freyd-scedrov-2}
   below.%  (where the variables in the axioms range over defined objects
   % only). However, in this reading of the axioms of Freyd and Scedrov
   % we had to (re-)introduce explicit strictness conditions to ensure
   % equivalence to the Axioms Set V by Scott.
 }

@@

Moreover, as an alternative to always unfolding the mapping from
$\FFOL$ to $\HOL$, abstract level proof tactics could be provided
e.g. in Isabelle/HOL to support intuitive interaction (and even
automation) in $\FFOL$ on top the semantical embedding.
